---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Hill Climbing as test for causal model

Project Description:

This project aims to build a predictive decision model that connects causal relationships between planting trees on farms and farmer livelihoods. The process will unfold in three main steps:

Defining the Problem and Building a Causal Model: The first step involves engaging experts and utilizing available literature to define the key causal relationships related to the intervention (e.g., planting trees on farms). A Directed Acyclic Graph (DAG) will be constructed to visually represent these relationships, incorporating various factors like costs, benefits, risks, and their impact on livelihoods.

Creating a Predictive Decision Model Using Expert Elicitation: In this phase, the causal model will be translated into a Bayesian Network (BN). Expert elicitation will be used to fill in the Conditional Probability Tables (CPTs) for each node in the BN, providing the probability distributions based on expert knowledge and available data.

Testing the Model with Hill Climbing Optimization: Once the Bayesian Network is constructed, data will be gathered (even if minimal) to test and refine the model. The hill climbing algorithm will be applied to optimize the model's parameters by adjusting them to best fit the observed data. The optimization process will focus on maximizing the model’s predictive accuracy and identifying the most likely causal relationships.

This approach integrates both qualitative insights from experts and quantitative data, allowing for a robust and adaptable decision model. The project will leverage causal inference techniques, expert-elicited probabilities, and optimization algorithms to improve decision-making for interventions aimed at enhancing livelihoods through agroforestry.

```{r plot_dagitty}
  source("dagitty_tree_planting.R")
# Plot the DAG
plot(dag)
```
We build the same graph in `bnlearn` for use in that environment. See all the custom CPTs in `model_in_bnlearn.R`. 

```{r plot_network_structure}
source("model_in_bnlearn.R")
plot(network_structure)
```

## Perform inference 

Calculate the probability of "Livelihoods" being "Improved" given "Trees on Farm". 

```{r inference_result}
inference_result <- cpquery(bn_fitted, event = (Livelihoods == "Improved"), evidence = (TreesOnFarm == "Yes"))

cat("Probability of improved livelihoods given trees on farm: ", inference_result, "\n")
```

To validate our Bayesian Network, we can perform several tests to ensure that the model behaves as expected and that the conditional dependencies between the nodes are correctly represented. 

### Inconsistent Evidence

Try introducing evidence that contradicts the CPTs and check for the system's response. For example, if a node is conditioned on one state, but the evidence conflicts with that, it should return a very low or zero probability.

```{r inconsistent_inference}
inconsistent_inference <- cpquery(bn_fitted, event = (TreesOnFarm == "No"), evidence = (Firewood == "Yes"))
print(inconsistent_inference)
```

This tests the model's behavior when evidence contradicts the dependency structure.

### Query for Node Probabilities

Test the probability distributions of individual nodes given various evidence. For example, given evidence for Market, check the conditional probability distribution for Livelihoods.

Example for Livelihoods:

```{r inference_Livelihoods}
inference_Livelihoods <- cpquery(bn_fitted, event = (Livelihoods == "Improved"), evidence = (Benefits == "High"))
print(inference_Livelihoods)
```

This should return the probability of improved livelihoods given that the market is high.

### Sensitivity Analysis

Perform a sensitivity analysis to understand how changes in one or more variables affect the results. For example, change the probability of Firewood or Timber and see how it affects the probability of Livelihoods.

```{r sensitivity_analysis}
sensitivity_analysis <- cpquery(bn_fitted, event = (Livelihoods == "Improved"), evidence = (Timber == "Yes"))
print(sensitivity_analysis)
```

### Simulation and Comparison with Expected Results

Generate synthetic data based on our network structure and compare it with expected or known results.

```{r simulated_data}
# Simulate 1000 samples
simulated_data <- rbn(bn_fitted, n = 1000)
head(simulated_data)
```

Calculate the observed distribution of 'Livelihoods'.

```{r observed_Livelihoods}
observed_Livelihoods <- table(simulated_data$Livelihoods) / nrow(simulated_data)
print(observed_Livelihoods)
```

Save the expectation for 'Livelihoods'. 

```{r expected_Livelihoods}
expected_Livelihoods <- c("Improved" = 0.7, "Not Improved" = 0.3)
```

Compare the observed distribution with the expected one. 

```{r observed_expected_compare}
data.frame(
  "Observed" = observed_Livelihoods,
  "Expected" = expected_Livelihoods
)
```

Calculate the distribution of 'Timber' given 'TreesOnFarm' (example for other node relationships too).

```{r}
observed_Timber_given_TreesOnFarm <- table(simulated_data$Timber, simulated_data$TreesOnFarm) / nrow(simulated_data)
print(observed_Timber_given_TreesOnFarm)
```

Visualize Livelihoods results. 

```{r}
library(ggplot2)

ggplot(simulated_data, aes(x = Livelihoods)) +
  geom_bar(aes(y = after_stat(prop)), stat = "count") +
  scale_y_continuous(labels = scales::percent) +
  ggtitle("Distribution of Livelihoods in Simulated Data")

```

### Hill-climbing algorithm

Learn the structure of a Bayesian network using a hill-climbing algorithm `hc`. 

```{r hill_climbing}
source("hill_climbing.R")
plot(fitted_model)
```

Using a hill-climbing algorithm to learn the structure of a Bayesian Network (BN) based on observed data. The observed data contains information from five publications, each contributing to various factors that may influence agroforestry systems and livelihoods in different regions. The goal is to use this data to infer the best network structure that best explains the dependencies among the variables in the dataset. See the details in `hill_climbing.R`.
